The Numerical Accuracy of Computing Electrostatic Interaction in the Inhomogeneous and Correlated Molecular Systems: for the Ewald Summation, SPME and Staggered Mesh Ewald Methods.

### 2012-07-11 longRange.hetero/alltext/publish/resubmit/main.tex
finish first revise, add figures

The calculation of the long-range electrostatic interaction is a very
important topic in the molecular simulations.  Due to the very slow
decay of the interaction strength with respect to the distance between charges,
the typical short-range fast algorithms (e.g. the cut-off + neighbor list
algorithm) do not work, and the
energy of the system is only conditionally convergent.  One of the
first algorithms handling this problem is the Ewald summation, dating
back to the 1920's~\cite{ewald1921die}.  It has been shown that
with a careful tuning of working parameters, the optimal computational
cost of the Ewald summation is  $\mathcal
O(N^{3/2})$~\cite{perram1988asc}, which is not feasible for modern
large scale molecular simulations.
Therefore, several Ewald based fast algorithms were proposed to reduce the
computational cost to a scalable level: $\mathcal O(N\log N)$. Some of them
are
the particle mesh Ewald (PME)
method~\cite{darden1993pme}, the smooth particle mesh Ewald (SPME)
method~\cite{essmann1995spm} and the particle-particle-particle-mesh
(P3M) method~\cite{hockney1988computer, deserno1998mue1}.
More recently, the staggered mesh or
``interlacing'' technique~\cite{chen1974reduction,eastwood1976optimal}
was applied to SPME
(called the staggered mesh Ewald)~\cite{cerutti2009staggered} and P3M
(called the interlaced P3M)~\cite{neelov2010interlaced}, and was showed to
improve the accuracy of these algorithms greatly.
In this paper, we focus on the Ewald summation, the SPME and
the staggered mesh Ewald methods. The Ewald summation is the starting
point of all the mentioned fast algorithm. The SPME
is the one of the most popular long-range algorithms among
the mainstream molecular simulation packages, for example,
AMBER~\cite{case2005amber}, GROMACS~\cite{van2005gromacs,
  hess2008gromacs} and NAMD~\cite{phillips2005scalable}.

The aforementioned fast algorithms have achieved great success
over the past three decades.
However, the fact that the user should provide six working parameters
for these methods may be problematic: it is difficult to try out the
most efficient parameter set in such a 6-dimensional parameter space.
The experiential choices may be either less accurate or wasting
computational effort.
This problem can be solved by an
in-depth study of the error introduced by these algorithms, namely the
error estimate~\cite{kolafa1992cutoff, hummer1995numerical,
  petersen1995accuracy, deserno1998mue2, stern2008mesh,
  wang2010optimizing, neelov2010interlaced, ballenegger2012convert},
by which the error is described by a function of the working
parameters.  Using the error estimates, a work flow has been designed to
automatically determine the nearly optimal combination of parameters,
in the sense that it is the fastest one satisfying a demanding
accuracy~\cite{wang2010optimizing}. However, so far, all error
estimates assume the homogeneity of the charge distribution, and the
independency of any pair of charges (i.e. the charges are
uncorrelated), but neither of the assumptions is satisfactorily
fulfilled in most molecular
systems of practical interest.  For example, the lipid bilayer membrane
solving in water: the hydrophobic tail(s) of the lipids carry no charge, and
the charges in the system
are correlated by the covalent bonds, the van der Waals
interaction, the hydrogen bonding network, etc..  Therefore, the current
error estimates may be problematic, or even not applicable.  Several
papers have shown that due to the correlation of charges, the error estimates
obviously deviate from the real
error~\cite{deserno1998mue2,wang2010optimizing,ballenegger2012convert}.

The purpose of this manuscript is to develop a reliable  error
estimate in inhomogeneous and correlated molecular systems.
We estimate the force error because the application is the molecular
dynamics (MD) simulation.  Firstly, a general error estimate
framework for the pairwise interaction is set up. We prove that the
error is decomposed into three additive parts: the homogeneity error,
the inhomogeneity error and the correlation error~\cite{wang2012}.
% These error estimates are only of theoretical interest, and
% not applicable, because the computational costs of the error estimates
% are $\mathcal O(N^2)$ (homogeneity and inhomogeneity error)
% and $\mathcal O(N^3)$ (correlation error).
% If the error force has a special form so that the error estimates
% can be written as convolutions, the computational costs
% are reduced to $\mathcal O(N\log N)$ and 
% $\mathcal O(N^2\log N)$, respectively.
The computational cost of the homogeneity and inhomogeneity
error is $\mathcal O(N\log N)$ when the error force kernel is
of convolution form.
The  correlation error can also be
estimated at a cost of $\mathcal O(N\log N)$
by the nearest neighbor approximation
technique, which \redc{assumes} the error  is dominated by  short-range correlation.
This
is the case for most the liquid and gas systems far from critical point.
Generally speaking,
when the error force can be written as a pairwise form, 
all estimate are derived easily under this framework.
Following this framework, we
develop the error estimates for the Ewald summation, the SPME 
and the staggered method Ewald methods.   The estimates
of the homogeneity and inhomogeneity errors are tested by two
randomly distributed inhomogeneous charges systems,
% (i.e. the charges are not correlated),
and is proved to
be very sharp.
The error estimates are also tested in a real inhomogeneous water system,
in which the charge correlation plays a very important role
on the force error. By 
including the nearest neighbor approximation of
\redc{the} correlation error, the quality of the estimates
is impressively enhanced.
This paper is ended by conclusions and remarks in
section~\ref{sec:conclusions}.

@@

\subsection{Nearest-neighbor approximation of the correlation

@@

So we consider the nearest-neighbor approximation to reduce

@@

strongly correlated. It could also be defined by the charges falling in some
certain neighboring range or charges connected by chemical bonds.

@@

As an example, we consider a typical three point charge water model
TIP3P~\cite{jorgensen1983comparison}: each hydrogen atom carries a
charge of $q_{\textrm{H}} = +0.417e$ and each oxygen atom carries a
charge of $q_{\textrm{O}} = -0.834e$. Within one molecule, the H-O
bond and H-O-H angle are rigid, so the correlation within one molecule
is considered important.  Therefore, in this case, we define the nearest neighbor by the
atoms within the same water molecule.  Denoting the index set of the oxygen
atoms by $\Omega_{\textrm{O}}$, and the index set of the hydrogen
atoms by $\Omega_{\textrm{H}}$, we have:

@@

which are accounting for the atomic correlation within one water

@@

% For the second term, do variable change $\v r'' = \v r' + \v s$, we have:
% \begin{align}
%   \textrm{II} = & 
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r')
%   \int_{\mathbb R^3}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_{q^2}(\v r', \v r' + \v s).
% \end{align}
% Split the integral region of the variable $\v s$ into the first
% neighbor shell $\mathcal B(\v r', R_s)$ and the reset $\mathbb R -
% \mathcal B(\v r', R_s)$. In the case of water, the first neighbor
% shell means the region where a single molecule occupies. Within this
% shell, the relative position of the oxygen and two hydrogens are
% always fixed. Outside of the first neighbor shell, the position of
% oxygens and hydrogens are belongs to different molecules, so the their
% positions are assumed to be independent. In most cases, the first
% neighbor shell is small compared to the simulation box size,
% therefore,
% \begin{align}\nonumber
%   \textrm{II}_{\textrm{out}} = & \,
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r')
%   \int_{\mathbb R^3 - \mathcal B(\v r', s)}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_{q^2}(\v r', \v r' + \v s) \\ \nonumber
%   = &\,
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r') \rho_q(\v r')
%   \int_{\mathbb R^3 - \mathcal B(\v r', s)}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_q(\v r' + \v s) \\
%   \approx &\,
%   q_i^2
%   [\v K\ast\rho_{q}]^2 (\v r_i)
% \end{align}

@@

The shift of the mesh can be effectively treated as shifting all
particles on the opposite direction. Since the system is periodic,
the direction of shift is not important. For convenience, we shift
all particles to the positive direction by half the mesh element size, namely
by a vector $\v s = \sum_\alpha\frac1{2K_\alpha}\v a_\alpha$. From
Eqn.~\eqref{eqn:error-A}, we have
\begin{align}
  \frac12\, [A(\v m, \v r) + A(\v m, \v r+\v s)]
  =&
  \sum_{\alpha=1}^3\sum_{l\neq 0}
  \frac12\,
  Z_{\alpha,l}(\v m)\,
  \big[
  (1 + e^{\pi i l})e^{2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r} - 2
  \big].
\end{align}
Therefore, when $l$ is odd, $1 + e^{\pi i l}$ vanishes. So the $\frac12\,[A(\v
m, \v r) + A(\v m, \v r+\v s)]$ can be approximated by the leading
order:
\begin{align}
  \frac12\, [A(\v m, \v r) + A(\v m, \v r+\v s)]
  \approx&
  -\sum_{\alpha=1}^3\sum_{l\neq 0}
  Z_{\alpha,l}(\v m).
\end{align}
Then from Eqn.~\eqref{eqn:ik.error.force}, we have the error force of
the staggered ik-differentiation:
\begin{align}\label{eqn:ik-st-error-force}
  \Delta\v F^{\textrm{ik,st}}_{\textrm{rec}}(\v r)
  \approx -2q
  \sum_jq_j
  \sum_\alpha\sum_{l\neq 0}
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  g(\v m) Z_{\alpha,l}(\v m)\,e^{2\pi i\v m\cdot(\v r-\v r_j)}.
\end{align}
Therefore, the error force kernel is
\begin{align}
  \v K^{\textrm{ik,st}}_{\textrm{rec}}(\v r)
  = -2
  \sum_\alpha\sum_{l\neq 0}\v G_{\alpha,l}(\v r),
\end{align}
then the corollary~\ref{thm:tmp2} is applicable.
% Moreover, since both the exact and the error
% force are anti-symmetric,  the Newton's third law is preserved by the
% calculated force.
We have the error estimates:
\begin{align}\label{eqn:ik.st.meanf}
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)
  \big\rangle
  & \approx
  - 2 q
  \big[
  \big(
  \sum_{\alpha}\sum_{l\neq 0}\v G_{\alpha,l}
  \big)
  \ast\rho_q\,
  \big] (\v r),
\end{align}
% Also neglect the high wave number terms,
and the RMS force error:
\begin{align}\label{eqn:ik.st.error}
  \big\langle
  \vert\Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)\vert^2
  \big\rangle
  \approx&\,
  4q^2\,
  \big[
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \big)^2
  \ast \rho_{q^2}\,
  \big] (\v r) +
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)
  \big\rangle^2.
\end{align}
The superscript ``st'' denotes the staggered mesh method.
The mean error force of the staggered ik-differentiation is the
same as the original ik-differentiation.
The RMS error is improved, because
$\big\langle\vert\Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r)\vert^2\big\rangle -
\big\langle\vert\Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)\vert^2\big\rangle=
2q^2
\big[\,
\big(
\sum_{\alpha} \sum_{l\neq 0}
\vert \v G_{\alpha,l}\vert^2
\big)
\ast \rho_{q^2}
\,\big] (\v r) \geq 0$.
The nearest neighbor approximation to the correlation error is:
\begin{align}\nonumber
  \mathcal E^{\textrm{ik,st}}_{\textrm{correlation}}(\v r)
  =\,&
  4q^2\,
  \big[
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \big)
  \cdot
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  [\,\hat T_{\textrm{O}}\hat{\v G}_{\alpha,l}]^{\vee}
  \big)
  \ast \rho_{\textrm{O}}\,
  \big] (\v r) \\\nonumber
  \,&+
  4q^2\,
  \big[
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \big)
  \cdot
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  [\,\hat T_{\textrm{H}}\hat{\v G}_{\alpha,l}]^{\vee}
  \big)
  \ast \rho_{\textrm{H}}\,
  \big] (\v r).
\end{align}

For the analytical differentiation, similarly we have: $\frac12\,[\v
B(\v m,\v r) + \v B(\v m, \v r+\v s)] \approx 0$. Therefore, from
Eqn.~\eqref{eqn:ana.error.force}, the error force of the staggered
analytical differentiation is approximately the same as the staggered
ik-differentiation, so are all the error estimates:
\begin{align}\label{eqn:ana-st-error-force}
  \Delta\v F^{\textrm{ana,st}}_{\textrm{rec}}(\v r)
  &\approx
  \Delta\v F^{\textrm{ik,st}}_{\textrm{rec}}(\v r)\\
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ana,st}}(\v r)
  \big\rangle
  &\approx
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)
  \big\rangle\\
  \big\langle
  \vert\Delta\v F_{\textrm{rec}}^{\textrm{ana,st}}(\v r)\vert^2
  \big\rangle
  &\approx
  \big\langle
  \vert\Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)\vert^2
  \big\rangle\\
  \mathcal E^{\textrm{ik,st}}_{\textrm{correlation}}(\v r)
  &\approx
  \mathcal E^{\textrm{ana,st}}_{\textrm{correlation}}(\v r)
\end{align}
The correction of the self-interacting error is not necessary,
because it is automatically canceled by the staggered mesh.
% The error from the self-interaction term is automatically fixed by the
% staggered mesh.
The RMS force error is improved because it is not
difficult to see 
$\big\langle\vert\Delta\v F_{\textrm{rec}}^{\textrm{ana}}(\v r)\vert^2\big\rangle -
\big\langle\vert\Delta\v F_{\textrm{rec}}^{\textrm{ana,st}}(\v r)\vert^2\big\rangle
\geq 0$.
Eqn.~\eqref{eqn:ana-st-error-force} implies, to the
leading order of approximation, the staggered
mesh ik- and analytical differentiation
are equivalent. Considering the latter uses
only one half of  FFTs as the former, it might be preferable
in large scale parallel simulations.
From Eqn.~\eqref{eqn:ik-st-error-force}
and~\eqref{eqn:ana-st-error-force}, it is obvious that the Newton's
third law is better preserved by the staggered mesh methods.

>>>>

The calculation of the long-range electrostatic interaction is a very
important topic in the molecular simulations.  Due to the very slow
decay of the interaction strength with respect to the distance between charges,
the typical short-range fast algorithms (e.g. the cut-off + neighbor list
algorithm) do not work, and the
energy of the system is only conditionally convergent.  One of the
first algorithms handling this problem is the Ewald summation, dating
back to the 1920's~\cite{ewald1921die}.  It has been shown that
with a careful tuning of working parameters, the optimal computational
cost of the Ewald summation is  $\mathcal
O(N^{3/2})$~\cite{perram1988asc}, which is not feasible for modern
large scale molecular simulations.
Therefore, several Ewald based fast algorithms were proposed to reduce the
computational cost to a scalable level: $\mathcal O(N\log N)$. Some of them
are
the particle mesh Ewald (PME)
method~\cite{darden1993pme}, the smooth particle mesh Ewald (SPME)
method~\cite{essmann1995spm} and the particle-particle-particle-mesh
(P3M) method~\cite{hockney1988computer, deserno1998mue1}.
\recheck{
It has been demonstrated that the SPME is actually
a special case of the P3M method~\cite{deserno1998mue1, ballenegger2012convert}.}
More recently, the staggered mesh or
``interlacing'' technique~\cite{chen1974reduction,eastwood1976optimal}
was applied to SPME
(called the staggered mesh Ewald)~\cite{cerutti2009staggered} and P3M
(called the interlaced P3M)~\cite{neelov2010interlaced}, and was showed to
improve the accuracy of these algorithms greatly.
In this paper, we focus on the Ewald summation, the SPME and
the staggered mesh Ewald methods. The Ewald summation is the starting
point of all the mentioned fast algorithm. The SPME
is the one of the most popular long-range algorithms among
the mainstream molecular simulation packages, for example,
AMBER~\cite{case2005amber}, GROMACS~\cite{van2005gromacs,
  hess2008gromacs} and NAMD~\cite{phillips2005scalable}.

The aforementioned fast algorithms have achieved great success
over the past three decades.
However, the fact that the user should provide six working parameters
for these methods may be problematic: it is difficult to find the
most efficient parameter set in such a 6-dimensional parameter space.
% The experiential choices may be either less accurate or wasting
% computational effort.
% This problem can be solved by an
\recheck{
  This problem can be solved by repeating and comparing
  the force calculation by
  using different parameter sets
  on a representative snapshot of the system,
  see e.g. Ref.~\cite{abraham2011optimization}.}
\recheck{An other approach is the 
  in-depth theoretical study}
of the error introduced by the algorithms, namely the
error estimate~\cite{kolafa1992cutoff, hummer1995numerical,
  petersen1995accuracy, deserno1998mue2, stern2008mesh,
  wang2010optimizing, neelov2010interlaced, ballenegger2012convert},
by which the error is described by a function of the working
parameters.  Using the error estimates, a work flow has been designed to
automatically determine the nearly optimal combination of parameters,
in the sense that it is the fastest one satisfying a demanding
accuracy~\cite{wang2010optimizing}. However, so far, all error
estimates assume the homogeneity of the charge distribution, and the
independency of any pair of charges (i.e. the charges are
uncorrelated), but neither of the assumptions is satisfactorily
fulfilled in most molecular
systems of practical interest.  For example, the lipid bilayer membrane
solving in water: the hydrophobic tail(s) of the lipids carry no charge, and
the charges in the system
are correlated by the covalent bonds, the van der Waals
interaction, the hydrogen bonding network, etc..  Therefore, the current
error estimates may be problematic, or even not applicable.  Several
papers have shown that due to the correlation of charges, the error estimates
obviously deviate from the real
error~\cite{deserno1998mue2,wang2010optimizing,ballenegger2012convert}.

The purpose of this manuscript is to develop a reliable  error
estimate in inhomogeneous and correlated molecular systems.
We estimate the force error because the application is the molecular
dynamics (MD) simulation.  Firstly, a general error estimate
framework for the pairwise interaction is set up.
\recheck{We show that the
  error may be approximately
  decomposed into three additive parts: the homogeneity error,
the inhomogeneity error and the correlation error~\cite{wang2012}.}
% These error estimates are only of theoretical interest, and
% not applicable, because the computational costs of the error estimates
% are $\mathcal O(N^2)$ (homogeneity and inhomogeneity error)
% and $\mathcal O(N^3)$ (correlation error).
% If the error force has a special form so that the error estimates
% can be written as convolutions, the computational costs
% are reduced to $\mathcal O(N\log N)$ and 
% $\mathcal O(N^2\log N)$, respectively.
The computational cost of the homogeneity and inhomogeneity
error is $\mathcal O(N\log N)$ when the error force kernel is
of convolution form.
The  correlation error can also be
estimated at a cost of $\mathcal O(N\log N)$
by the nearest neighbor approximation
technique, which \redc{assumes} the error  is dominated by  short-range correlation.
This
is the case for most the liquid and gas systems far from critical point.
Generally speaking,
when the error force can be written as a pairwise form, 
all estimate are derived easily under this framework.
Following this framework, we
develop the error estimates for the Ewald summation, the SPME 
and the staggered method Ewald methods.   The estimates
of the homogeneity and inhomogeneity errors are tested by two
randomly distributed inhomogeneous charges systems,
% (i.e. the charges are not correlated),
and is proved to
be very sharp.
The error estimates are also tested in a real inhomogeneous water system,
in which the charge correlation plays a very important role
on the force error. By 
including the nearest neighbor approximation of
\redc{the} correlation error, the quality of the estimates
is impressively enhanced.
This paper is ended by conclusions and remarks in
section~\ref{sec:conclusions}.

@@

\subsection{Nearest neighbor approximation of the correlation

@@

So we consider the nearest neighbor approximation to reduce

@@

strongly correlated. It could be defined \recheck{either}
by the charges falling in some
certain neighboring range, or \recheck{by} charges connected by chemical bonds.

@@

As an example, we consider a typical three point charge water model
TIP3P~\cite{jorgensen1983comparison}: each hydrogen atom carries a
charge of $q_{\textrm{H}} = +0.417e$ and each oxygen atom carries a
charge of $q_{\textrm{O}} = -0.834e$.  \recheck{ There are two types
  of correlation in this system: (1) the bonded correlation
  that originates from the rigid H-O
  bond and H-O-H angle within one molecule,
  and (2) the
  non-bonded correlation or the
  intermolecular correlation that originates from
  a complex coupling of various intermolecular interaction, such as
  the van der Waals interaction,
  the hydrogen bonding and etc..
% Within one molecule, the H-O bond and
% H-O-H angle are rigid, so the correlation within one molecule is
% considered important.
% Therefore, in this case,
}

We firstly consider the bonded correlation, and define the nearest
neighbor by the atoms within the same water molecule.  Denoting the
index set of the oxygen atoms by $\Omega_{\textrm{O}}$, and the index
set of the hydrogen atoms by $\Omega_{\textrm{H}}$, we have:

@@

which are accounting for the bonded correlation within one water

@@

\recheck{
It should be noted that we define the force error by inserting a
testing charge into the water system, and comparing the
calculated and exact force on it. This definition implicitly assumes
the independency between the testing particle and the
system, as illustrated by Fig.~\ref{fig:tmp-add0}.
% Therefore, the correlation error
% estimate~\eqref{eqn:c-error-water}
% actually neglects the correlation
% of the testing particle to the system, and
% calculates the error from all bonded correlations
% within its {neighboring} molecules, as illustrated by Fig.~\ref{fig:tmp-add0}.
It is consistent
with the usual setting of the bonded pair atoms: the electrostatic interaction
between them is considered to be \emph{not} calculated.
Throughout this paper, we remove this  interaction from the
standard long-range algorithms.}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{correlation.1.eps}
  \caption{The schematic plot of two interacting water molecules.
    When we calculate the force error of the left oxygen atom (it is now
    a testing charge), the
    correlation with its bonded hydrogens are not considered, but
    the correlations within any neighboring molecule are considered
    by the bonded nearest neighbor approximation,
    as indicated by the solid lines with two arrows.}
  \label{fig:tmp-add0}
\end{figure}

\recheck{
  For the non-bonded correlation,
  we define the nearest neighbor
  by non-bonded atoms closer than a certain radius $R_g$ (neighboring range).
  If the system is homogeneous and isotropic, the pair
  correlation function $\rho^{(2)}(\v r', \v r'')$ in
  Eqn.~\eqref{eqn:start-corr} can be represented by the radial
  distribution functions (RDFs).
  Following the same way as the bonded correlation,
  the non-bonded correlation error can be estimated:
  the functions $T_{\textrm{O}}$ and $T_{\textrm{H}}$
  are given by
  \begin{align}\label{eqn:nb-nna-1}
    \hat T_{\textrm{O}} (\v m)
    &= \int_0^{R_g}
    [q_{_{\textrm{H}}} \rho_{_{\textrm{H}}} \tilde g_{_{\textrm{OH}}}(r)
    +q_{_{\textrm{O}}} \rho_{_{\textrm{O}}} \tilde g_{_{\textrm{OO}}}(r)]
    \,\frac{\sin(2\pi m r)}{2\pi mr}\,4\pi r^2 \d dr,\\\label{eqn:nb-nna-2}
    \hat T_{\textrm{H}} (\v m)
    &= \int_0^{R_g}
    [q_{_{\textrm{H}}} \rho_{_{\textrm{H}}} \tilde g_{_{\textrm{HH}}}(r)
    +q_{_{\textrm{O}}} \rho_{_{\textrm{O}}} \tilde g_{_{\textrm{OH}}}(r)]
    \,\frac{\sin(2\pi m r)}{2\pi mr}\,4\pi r^2 \d dr,
  \end{align}
  where
  $\tilde g_{_{\textrm{OO}}}(r) = g_{_{\textrm{OO}}}(r) - 1$,
  $\tilde g_{_{\textrm{OH}}}(r) = g_{_{\textrm{OH}}}(r) - 1$ and
  $\tilde g_{_{\textrm{HH}}}(r) = g_{_{\textrm{HH}}}(r) - 1$.
  $g_{_{\textrm{OO}}}(r)$, $g_{_{\textrm{OH}}}(r)$ and $g_{_{\textrm{HH}}}(r)$
  are the RDFs of oxygen-oxygen, oxygen-hydrogen and hydrogen-hydrogen,
  respectively. $\rho_{_{\textrm{H}}}$ and $\rho_{_{\textrm{O}}}$ are
  the homogeneous number densities
  of the hydrogen and oxygen atoms, respectively.
  For a short-range correlated system, the RDFs converges to 1
  as $r$ goes to infinity, so the integrals in Eqn.~\eqref{eqn:nb-nna-1}
  and \eqref{eqn:nb-nna-2} are convergent with respect to
  increasing $R_g$.

  Both the bonded and non-bonded nearest neighbor approximations
  require some uniformity of the correlation in the system:
  the bonded estimate requires the uniformity of the water
  direction preference, while the non-bonded estimate requires the uniformity
  of the density distribution and the isotropicity of the system.
  These requirements should be carefully checked before using
  these error estimates.
  If the they are seriously violated,
  one can either use these estimates for 
  some regions (or sub-systems of the whole system) that
  still preserve the uniformity,
  or estimate the correlation error at the cost of
  $\mathcal O(N^2\log N)$, if the correlation in the system is highly
  inhomogeneous.  
  % one can either estimate the correlation error at the cost of
  % $\mathcal O(N^2\log N)$, or estimate the correlation error for
  % some regions (or sub-system of the whole system) that
  % still preserve those uniformity.

}
% For the second term, do variable change $\v r'' = \v r' + \v s$, we have:
% \begin{align}
%   \textrm{II} = & 
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r')
%   \int_{\mathbb R^3}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_{q^2}(\v r', \v r' + \v s).
% \end{align}
% Split the integral region of the variable $\v s$ into the first
% neighbor shell $\mathcal B(\v r', R_s)$ and the reset $\mathbb R -
% \mathcal B(\v r', R_s)$. In the case of water, the first neighbor
% shell means the region where a single molecule occupies. Within this
% shell, the relative position of the oxygen and two hydrogens are
% always fixed. Outside of the first neighbor shell, the position of
% oxygens and hydrogens are belongs to different molecules, so the their
% positions are assumed to be independent. In most cases, the first
% neighbor shell is small compared to the simulation box size,
% therefore,
% \begin{align}\nonumber
%   \textrm{II}_{\textrm{out}} = & \,
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r')
%   \int_{\mathbb R^3 - \mathcal B(\v r', s)}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_{q^2}(\v r', \v r' + \v s) \\ \nonumber
%   = &\,
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r') \rho_q(\v r')
%   \int_{\mathbb R^3 - \mathcal B(\v r', s)}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_q(\v r' + \v s) \\
%   \approx &\,
%   q_i^2
%   [\v K\ast\rho_{q}]^2 (\v r_i)
% \end{align}

@@

The shift of the mesh can be effectively treated as shifting all
particles on the opposite direction. Since the system is periodic,
the direction of shift is not important. For convenience, we shift
all particles to the positive direction by half the mesh element size, namely
by a vector $\v s = \sum_\alpha\frac1{2K_\alpha}\v a_\alpha$. From
Eqn.~\eqref{eqn:error-A}, we have
\begin{align}
  \frac12\, [A(\v m, \v r) + A(\v m, \v r+\v s)]
  =&
  \sum_{\alpha=1}^3\sum_{l\neq 0}
  \frac12\,
  Z_{\alpha,l}(\v m)\,
  \big[
  (1 + e^{\pi i l})e^{2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r} - 2
  \big].
\end{align}
Therefore, when $l$ is odd, $1 + e^{\pi i l}$ vanishes. So the $\frac12\,[A(\v
m, \v r) + A(\v m, \v r+\v s)]$ can be approximated by the leading
order:
\begin{align}\label{eqn:leading-order-1}
  \frac12\, [A(\v m, \v r) + A(\v m, \v r+\v s)]
  \approx&
  -\sum_{\alpha=1}^3\sum_{l\neq 0}
  Z_{\alpha,l}(\v m).
\end{align}
\recheck{The validity of this approximation and a similar approximation
  \eqref{eqn:leading-order-2}
  will be numerically checked in Sec.~\ref{sec:example3}.}
Then from Eqn.~\eqref{eqn:ik.error.force}, we have the error force of
the staggered ik-differentiation:
\begin{align}\label{eqn:ik-st-error-force}
  \Delta\v F^{\textrm{ik,st}}_{\textrm{rec}}(\v r)
  \approx -2q
  \sum_jq_j
  \sum_\alpha\sum_{l\neq 0}
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  g(\v m) Z_{\alpha,l}(\v m)\,e^{2\pi i\v m\cdot(\v r-\v r_j)}.
\end{align}
Therefore, the error force kernel is
\begin{align}
  \v K^{\textrm{ik,st}}_{\textrm{rec}}(\v r)
  = -2
  \sum_\alpha\sum_{l\neq 0}\v G_{\alpha,l}(\v r),
\end{align}
then the corollary~\ref{thm:tmp2} is applicable.
% Moreover, since both the exact and the error
% force are anti-symmetric,  the Newton's third law is preserved by the
% calculated force.
We have the error estimates:
\begin{align}\label{eqn:ik.st.meanf}
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)
  \big\rangle
  & \approx
  - 2 q
  \big[
  \big(
  \sum_{\alpha}\sum_{l\neq 0}\v G_{\alpha,l}
  \big)
  \ast\rho_q\,
  \big] (\v r),
\end{align}
% Also neglect the high wave number terms,
and the RMS force error:
\begin{align}\label{eqn:ik.st.error}
  \big\langle
  \vert\Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)\vert^2
  \big\rangle
  \approx&\,
  4q^2\,
  \big[
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \big)^2
  \ast \rho_{q^2}\,
  \big] (\v r) +
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)
  \big\rangle^2.
\end{align}
The superscript ``st'' denotes the staggered mesh method.
The mean error force of the staggered ik-differentiation is the
same as the original ik-differentiation.
The RMS error is improved, because
$\big\langle\vert\Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r)\vert^2\big\rangle -
\big\langle\vert\Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)\vert^2\big\rangle=
2q^2
\big[\,
\big(
\sum_{\alpha} \sum_{l\neq 0}
\vert \v G_{\alpha,l}\vert^2
\big)
\ast \rho_{q^2}
\,\big] (\v r) \geq 0$.
The nearest neighbor approximation to the correlation error is:
\begin{align}\nonumber
  \mathcal E^{\textrm{ik,st}}_{\textrm{correlation}}(\v r)
  =\,&
  4q^2\,
  \big[
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \big)
  \cdot
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  [\,\hat T_{\textrm{O}}\hat{\v G}_{\alpha,l}]^{\vee}
  \big)
  \ast \rho_{\textrm{O}}\,
  \big] (\v r) \\\nonumber
  \,&+
  4q^2\,
  \big[
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \big)
  \cdot
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  [\,\hat T_{\textrm{H}}\hat{\v G}_{\alpha,l}]^{\vee}
  \big)
  \ast \rho_{\textrm{H}}\,
  \big] (\v r).
\end{align}
